### 2024-08-07 paper/paper.qmd
Data commit

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Rahma Binth Mohammad
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)

total_ex_cpi <- read_csv(here::here("data/analysis_data/total_ex_cpi.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}
```{r}
#| fig-cap: Linear model
#| echo: false
#| warning: false
#| message: false

total_ex_cpi |>
  ggplot(aes(x = cpi, y = expenditure, color = factor(Year))) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.5, aes(group = 1)) + 
  labs(
    title = "Total Expenditure vs All-Items CPI",
    x = "All-Items CPI",
    y = "Total Expenditure",
    color = "Year"
  ) +
  theme_minimal() +
  scale_color_viridis_d()
```

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

```

Talk more about it.

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

### Simple linear regression

Define $y_i$ as the total expenditure in Canadian dollars in year i. Then $cpi_i$ is the annual Consumer Price Index (CPI) for year i

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{cpi}_i + \epsilon\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

Where

### Multiple linear regression

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{cpi}_i + \beta_2 \mbox{income}_i+ \epsilon\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

```{r}
#| label: fig-2
#| fig-cap: Linear model
#| echo: false
#| warning: false
#| message: false
library(scales)

total_ex_cpi |>
  ggplot(aes(x = cpi, y = expenditure, color = factor(Year))) +
  geom_point(size = 1) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  labs(
    title = "Total Expenditure vs All-Items CPI",
    x = "All-Items CPI",
    y = "Total Expenditure",
    color = "Year"
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_viridis_d() +
  theme_minimal()
```

```{r}
#| label: fig-3
#| fig-cap: Linear model
#| echo: false
#| warning: false
#| message: false
library(scales)

total_ex_cpi |>
  ggplot(aes(x = cpi, y = expenditure, color = factor(Year))) +
  geom_point(size = 1) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  labs(
    title = "Total Expenditure vs All-Items CPI",
    x = "All-Items CPI",
    y = "Total Expenditure",
    color = "Year"
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_viridis_d() +
  theme_minimal()
```
# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
second_model <-
  readRDS(file = here::here("models/second_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of "
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model,
    "Second model" = second_model
  ),
  statistic = "mad",
  fmt = 2
)
```

```{r}

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

### 2024-08-07 scripts/00-install_packages.R

@@ -1,0 +1,12 @@
+#### Preamble ####
+# Purpose: Installs the necessary packages. 
+# Author: Rahma Binth Mohammad
+# Date: 
+# Contact: rahma.binthmohammad@mail.utoronto.ca
+# License: MIT
+
+#### Download packages ####
+install.packages("tidyverse")
+install.packages("cansim")
+install.packages("dplyr")
+

### 2024-08-07 scripts/01-download_data.R

@@ -1,0 +1,28 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from Statistics Canada
+# Author: Rahma Binth Mohammad
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rahma.binthmohammad@mail.utoronto.ca
+# Pre-requisites: Run "00-install_packages.R".
+# Any other information needed? [...UPDATE THIS...]
+
+#### Workspace setup ####
+library(cansim)
+library(tidyverse)
+library(dplyr)
+
+#### Download data ####
+
+#Household spending data
+spending <- get_cansim("11-10-0222-01")
+
+#Annual CPI data
+cpi <- get_cansim("18-10-0005-01")
+
+#Household income
+income <- get_cansim("11-10-0008-01")
+
+#### Save data ###
+write_csv(spending, "data/raw_data/spending.csv") 
+write_csv(cpi, "data/raw_data/cpi.csv") 
+write_csv(income, "data/raw_data/income.csv") 

### 2024-08-07 scripts/02-data_cleaning.R

@@ -1,0 +1,128 @@
+#### Preamble ####
+# Purpose: Cleans the raw plane data recorded by Statistics Canada and filters the necessary columns
+# Author: Rahma Binth Mohammad
+# Date: 6 April 2023 [...UPDATE THIS...]
+# Contact: rahma.binthmohammad@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run "00-install_packages.R" and "01-download_data". 
+# Any other information needed? [...UPDATE THIS...]
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Clean data ####
+
+#Filter and rename the household spending data
+spending <- read.csv("data/raw_data/spending.csv")
+
+spending <- spending %>% 
+  rename(expenditure_categories = Household.expenditures..summary.level.categories, Year = REF_DATE)
+
+spending <- spending %>%
+  filter(GEO == "Ontario",
+         expenditure_categories %in% c("Total expenditure", "Shelter", "Food expenditures", "Recreation")) %>%
+  select(Year, GEO, expenditure_categories, VALUE)
+
+#Filter and rename the annual CPI data
+cpi <- read.csv("data/raw_data/cpi.csv")
+
+cpi <- cpi %>% 
+  rename(product_groups = Products.and.product.groups, Year = REF_DATE)
+  
+cpi <- cpi %>%
+  filter(GEO == "Ontario", Year >= 2010, Year <= 2021,
+         product_groups %in% c("All-items", "Shelter", "Food", "Recreation, education and reading")) %>%
+  select(Year, GEO, product_groups, VALUE)
+
+#Filter and rename the household income data
+income <- read.csv("data/raw_data/income.csv")
+
+income <- income %>% 
+  rename(income_categories = Persons.with.income, Year = REF_DATE, age = Age.group)
+
+income <- income %>%
+  filter(GEO == "Ontario", Year >= 2010, Year <= 2021, age == "All age groups",
+         Sex == "Both sexes", income_categories == "All persons with income") %>%
+  select(Year, GEO, Sex, income_categories, VALUE)
+
+#Combine household spending and annual CPI data
+spending_n_cpi <- bind_rows(spending, cpi)
+
+#Isolate total household spending and all-items CPI data with just year and value
+total_ex_year <- spending %>%
[…]

### 2024-08-07 scripts/03-test_data.R

@@ -1,0 +1,47 @@
+#### Preamble ####
+# Purpose: Tests... [...UPDATE THIS...]
+# Author: Rohan Alexander [...UPDATE THIS...]
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# License: MIT
+# Pre-requisites: Run the "00-install_packages.R", "01-download_data", and "02-data_cleaning"
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Load data ####
+spending <- read.csv("data/analysis_data/spending.csv")
+cpi <- read.csv("data/analysis_data/cpi.csv")
+income <- read.csv("data/analysis_data/income.csv")
+
+#### Test data ####
+#Check class of data columns for all three datasets
+spending$Year |> class() == "integer"
+spending$GEO |> class() == "character"
+spending$expenditure_categories |> class() == "character"
+spending$VALUE |> class() == "integer"
+
+cpi$Year |> class() == "integer"
+cpi$GEO |> class() == "character"
+cpi$product_groups |> class() == "character"
+cpi$VALUE |> class() == "numeric"
+
+income$Year |> class() == "integer"
+income$GEO |> class() == "character"
+income$Sex |> class() == "character"
+income$income_categories |> class() == "character"
+income$VALUE |> class() == "integer"
+
+#Check years in common
+years_spending <- unique(spending$Year)
+years_cpi <- unique(cpi$Year)
+years_income <- unique(income$Year)
+years_common <- Reduce(intersect, list(years_spending, years_cpi, years_income))
+
+#Check missing values
+is.na(spending)
+is.na(cpi)
+is.na(income)
+

### 2024-08-07 scripts/04-model.R

@@ -1,0 +1,51 @@
+#### Preamble ####
+# Purpose: Models... [...UPDATE THIS...]
+# Author: Rahma Binth Mohammad
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rahma.binthmohammad@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Run the "00-install_packages.R", "01-download_data.R", "02-data_cleaning.R", "03-test_data.R"
+# Any other information needed? [...UPDATE THIS...]
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+#### Read data ####
+total_ex_cpi <- read.csv("data/analysis_data/total_ex_cpi.csv") 
+ex_cpi_income <- read.csv("data/analysis_data/ex_cpi_income.csv")
+
+### Model data ####
+first_model <-
+  stan_glm(
+    formula = expenditure ~ cpi,
+    data = total_ex_cpi,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 853
+  )
+
+second_model <-
+  stan_glm(
+    formula = expenditure ~ cpi + income,
+    data = ex_cpi_income,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 853
+  )
+  
+#### Save model ####
+saveRDS(
+  first_model,
+  file = "models/first_model.rds"
+)
+
+saveRDS(
+  second_model,
+  file = "models/second_model.rds"
+)
+

### 2024-08-07 scripts/05-graphs.R

@@ -1,0 +1,113 @@
+#### Preamble ####
+# Purpose: Creates graphs with filtered data
+# Author: Rahma Binth Mohammad
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rahma.binthmohammad@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+#### Workspace setup ####
+library(tidyverse)
+library(dplyr)
+
+#### Load data ####
+spending <- read.csv("data/analysis_data/spending.csv")
+
+summary(spending)
+# Calculate summary statistics
+spending_stats <- spending %>%
+  group_by(expenditure_categories) %>%
+  summarise(
+    count = n(),
+    mean_value = mean(VALUE, na.rm = TRUE),
+    median_value = median(VALUE, na.rm = TRUE),
+    sd_value = sd(VALUE, na.rm = TRUE),
+    min_value = min(VALUE, na.rm = TRUE),
+    max_value = max(VALUE, na.rm = TRUE),
+    .groups = 'drop'
+  )
+
+# Print summary statistics
+print(spending_stats)
+
+#### Load data ####
+cpi <- read.csv("data/analysis_data/cpi.csv")
+
+# Calculate summary statistics
+cpi_stats <- cpi %>%
+  group_by(product_groups) %>%
+  summarise(
+    count = n(),
+    mean_value = mean(VALUE, na.rm = TRUE),
+    median_value = median(VALUE, na.rm = TRUE),
+    sd_value = sd(VALUE, na.rm = TRUE),
+    min_value = min(VALUE, na.rm = TRUE),
+    max_value = max(VALUE, na.rm = TRUE),
+    .groups = 'drop'
+  )
+
+# Print summary statistics
+print(cpi_stats)
+
+#### Load data ####
+income <- read.csv("data/analysis_data/income.csv")
+
+# Calculate and print summary statistics
+# Only one category in data - "All persons with income"
+summary(income$VALUE)
+
+ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5, aes(group = 1)) + # Connect the points with a line
+  labs(
+    title = "Total Expenditure vs All-Items CPI",
+    x = "All-Items CPI",
[…]

### 2024-08-08 paper/paper.qmd
Update graphs and add to paper

@@
   - Rahma Binth Mohammad
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/rahmabinth/consumer-behaviour"
 date: today
@@
 library(palmerpenguins)
+library(patchwork)
+spending <-  read_csv(here::here("data/analysis_data/spending.csv"))
+cpi <-  read_csv(here::here("data/analysis_data/cpi.csv"))
+income <- read_csv(here::here("data/analysis_data/income.csv"))
 
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+# Data {#sec-data}
 
-The remainder of this paper is structured as follows. @sec-data....
+```{r}
+#| label: fig-spending
+#| fig-cap: Household spending per year
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+#Plot spending categories for the years
+spending_colours <- c(
+  "Total expenditure" = "goldenrod3", 
+  "Food" = "purple",
+  "Shelter" = "green",
+  "Recreation" = "blue")
+ggplot(spending, aes(x = Year, y = VALUE, color = expenditure_categories, group = expenditure_categories)) +
+  geom_line() +
+  geom_point() +
+  labs(
 
-# Data {#sec-data}
+    x = "Year",
+    y = "Expenditure Value ($)"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = spending_colours, name = NULL) +
+  scale_x_continuous(
+    breaks = seq(2010, 2021, by = 2)) +
+  scale_y_continuous(
+      labels = scales::comma
+    )
+```
+From the spending dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
+This choice was made because these are principal expenditures for everyone.
+Recreation was added to this mix, to identify how consumers behave when inflation is high.
+
 ```{r}
-#| fig-cap: Linear model
+#| label: fig-cpi
+#| fig-cap: Consumer Price Index (CPI) per Year
+#| fig-align: "center"
 #| echo: false
@@
 
-total_ex_cpi |>
-  ggplot(aes(x = cpi, y = expenditure, color = factor(Year))) +
+cpi_colours <- c(
+  "All-items" = "goldenrod3", 
+  "Food" = "purple",
+  "Shelter" = "green",
+  "Recreation" = "blue")
+
+ggplot(cpi, aes(x = Year, y = VALUE, color = product_groups, group = product_groups)) +
+  geom_line() +
+  geom_point() +
+  labs(
+    x = "Year",
+    y = "CPI",
+  ) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = cpi_colours, name = NULL) +
+  scale_x_continuous(
+    breaks = seq(2010, 2021, by = 2)) +
+  scale_y_continuous(
+    breaks = seq(100, 160, by = 10),
+    limits = c(100, 160)
+  )
+
+```
+
+From the CPI dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures. The variables
+
+```{r}
+#| label: fig-income
+#| fig-cap: Number of Income Earners
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+#Plot the number of income earners against the year
+ggplot(income, aes(x = Year, y = VALUE)) +
+  geom_line(colour = "darkred") +
+  geom_point(colour = "darkred") +
+  labs(
+    x = "Year",
+    y = "Income Earners"
+  ) +
+  theme_minimal() +
+  scale_x_continuous(
+    breaks = seq(2010, 2021, by = 2)) +
+  scale_y_continuous(
+    labels = scales::comma,
+    limits = c(9600000, 11250000)
+  )
+
+```
+
+From the income dataset, we chose to isolate the values for the category “all persons with income”.
+This indicate all tax fillers and dependants that earn income from all sources.
+This was done to keep our analysis consistent.
+
+```{r}
+#| label: fig-total_ex_cpi
+#| fig-cap: CPI and total expenditure over the year
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
   geom_point(size = 1) +
-  geom_line(linewidth = 0.5, aes(group = 1)) + 
+  geom_line(linewidth = 0.5, aes(group = 1)) + # Connect the points with a line
   labs(
-    title = "Total Expenditure vs All-Items CPI",
     x = "All-Items CPI",
@@
   theme_minimal() +
-  scale_color_viridis_d()
+  scale_color_viridis_d() +
+  scale_x_continuous(
+    limits = c(115, 145)) +
+  scale_y_continuous(
+    labels = scales::comma,
+    limits = c(75000, 100000)
+  )
 ```
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-```
-
-Talk more about it.
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-```
-
-Talk way more about it.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+The goal of our model is to forecasting the total expenditure in a household in a year, based solely on a the Consumer Price Index, and the number of people in Ontario earning an income.
+We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
+We use the default priors from `rstanarm`.
 
@@
 ### Simple linear regression
-
-Define $y_i$ as the total expenditure in Canadian dollars in year i. Then $cpi_i$ is the annual Consumer Price Index (CPI) for year i
+The model we employ is:
 
@@
 
-Where
+Where:
+
+* $y_i$ as the total expenditure in Canadian dollars in year i. 
+* $cpi_i$ is the annual Consumer Price Index (CPI) for year i
 
 ### Multiple linear regression
+The model we employ is:
 
@@
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+Where:
+* $y_i$ as the total expenditure in Canadian dollars in year i. 
+* $cpi_i$ is the annual Consumer Price Index (CPI) for year i
+* $income_i$ is the number of people in Ontario earning an income
 

### 2024-08-08 scripts/01-download_data.R

@@ -19,10 +19,10 @@ spending <- get_cansim("11-10-0222-01")
 #Annual CPI data
 cpi <- get_cansim("18-10-0005-01")
 
-#Household income
-income <- get_cansim("11-10-0008-01")
+#Household income, Data is too large to push to git, hence it is downloaded. You can use this code to get it directly.
+#income <- get_cansim("11-10-0008-01")
 
 #### Save data ###
 write_csv(spending, "data/raw_data/spending.csv") 
 write_csv(cpi, "data/raw_data/cpi.csv") 
-write_csv(income, "data/raw_data/income.csv") 
+#write_csv(income, "data/raw_data/income.csv") 

### 2024-08-08 scripts/02-data_cleaning.R

@@ -23,6 +23,12 @@ spending <- spending %>%
          expenditure_categories %in% c("Total expenditure", "Shelter", "Food expenditures", "Recreation")) %>%
   select(Year, GEO, expenditure_categories, VALUE)
 
+spending <- spending %>%
+  mutate(expenditure_categories = recode(
+    expenditure_categories,
+    "Food expenditures" = "Food"
+  ))
+
 #Filter and rename the annual CPI data
 cpi <- read.csv("data/raw_data/cpi.csv")
 
@@ -33,6 +39,12 @@ cpi <- cpi %>%
   filter(GEO == "Ontario", Year >= 2010, Year <= 2021,
          product_groups %in% c("All-items", "Shelter", "Food", "Recreation, education and reading")) %>%
   select(Year, GEO, product_groups, VALUE)
+
+cpi <- cpi %>%
+  mutate(product_groups = recode(
+    product_groups,
+    "Recreation, education and reading" = "Recreation"
+  ))
 
 #Filter and rename the household income data
 income <- read.csv("data/raw_data/income.csv")

### 2024-08-08 scripts/05-graphs.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Creates graphs with filtered data
+# Purpose: Creates graphs with filtered data that is included in the paper, useful for reproduction
 # Author: Rahma Binth Mohammad
 # Date: 11 February 2023 [...UPDATE THIS...]
 # Contact: rahma.binthmohammad@mail.utoronto.ca
@@ -10,6 +10,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+library(scales)
 
 #### Load data ####
 spending <- read.csv("data/analysis_data/spending.csv")
@@ -57,17 +58,98 @@ income <- read.csv("data/analysis_data/income.csv")
 # Only one category in data - "All persons with income"
 summary(income$VALUE)
 
+#Plot spending categories for the years
+spending_colours <- c(
+  "Total expenditure" = "goldenrod3", 
+  "Food" = "purple",
+  "Shelter" = "green",
+  "Recreation" = "blue")
+spending_n_year <- ggplot(spending, aes(x = Year, y = VALUE, color = expenditure_categories, group = expenditure_categories)) +
+  geom_line() +
+  geom_point() +
+  labs(
+    title = "Household Spending per Year",
+    x = "Year",
+    y = "Expenditure Value ($)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    legend.position = "bottom") +
+  scale_color_manual(values = spending_colours, name = NULL) +
+  scale_x_continuous(
+    breaks = seq(2010, 2021, by = 2)) +
+  scale_y_continuous(
+      labels = scales::comma
+    )
+
+print(spending_n_year)
+#Plot CPI values for each category for each year
+#Set colours so that it is easily comparible with spending data
+cpi_colours <- c(
+  "All-items" = "goldenrod3", 
+  "Food" = "purple",
+  "Shelter" = "green",
+  "Recreation" = "blue")
+
+cpi_n_year <- ggplot(cpi, aes(x = Year, y = VALUE, color = product_groups, group = product_groups)) +
+  geom_line() +
+  geom_point() +
+  labs(
+    title = "Consumer Price Index (CPI) per Year",
+    x = "Year",
+    y = "CPI",
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5), 
[…]

### 2024-08-08 paper/paper.qmd
Update graphs for data section

@@
 # Data {#sec-data}
+## Sources
+A good indicator of inflation is the Consumer Price Index (CPI), as it measures the change in price using a fixed basket of goods and services.
+Statistics Canada is a government agency that collects information on Canada’s economy (cite).
+Thus, it has the data that provides the CPI for numerous years.
+Statistics Canada also provides us with other datasets such as the average household income and the average household expenditures, in a year for all the provinces.
+We have chosen to look at Ontario solely, as we hold firmly Ontario behaves very differently due to its high population in relation to the other provinces.
+The data for Canada as a whole is complicated as there are many missing values for some provinces.
+Our analysis and the data provided by Statistics Canada makes it possible to reproduce our model for different provinces using the github repository.
+While there are other datasets available, these three datasets are provided by the same source and has data for provinces across Canada.
+The expenditure and cpi data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question.
+@sec-variables will illustrate the visualised data.
+
+This paper’s objective is to address the following questions using data collected from a government census: (1) Does changes in prices affect expenditure?
+(2) Does changed in prices affect expenditure as income changes?
+
+### Spending Data
+
+`spending.csv`
+
+This dataset contains the average total expenditures of the Ontario resident for the years 2010 to 2021, excluding years 2018 and 2020.
+It also contains the breakdown of the expenditures, food purchased (from restaurant and stores), shelter (rented and owned), utilities such as water, fuel, and electricity for the principal accommodation, clothing, transportation (private and public), healthcare, recreational goods and services.
+
+### Inflation Data
+
+`cpi.csv`
+
+The dataset provides CPI for every year.
+It is not seasonally adjusted for changes as it is an annual average.
+Thus, it is actual price paid by consumers and not adjusted for seasonal fluctuations.
+The base year for the CPI calculation is 2002 where CPI is 100.
+Thus, if our CPI is 120 for a year, it indicates a 20% increase since 2002.
+This value is useful as an indicator for inflation, as it provides the increase and comparison between years, rather than a absolute price values.
+
+### Income Data
+
+`income.csv`
+
+Throughout the province of Ontario, there exits people in many different financial situations.
+Statistics Canada provides us with this dataset that contains the average number of people in an income category.
+For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Please note as for 2020, COVID-19 government income support and benefits are included in income estimates.
+
+## Methodology
+The language and environment used for this analysis is R (R Core Team 2020), alongside the tidyverse (cite), rstanarm (cite), cansim (cite), knitr (cite), ggplot2 (cite), scales (cite), modelsummary (cite), and marginaleffects (cite) packages.
+
+We downloaded the data directly from Statistics Canada using the cansim package. We cleaned it, identified missing values and wrote csv files to make it easily accessible using R studio.
+
+## Variables {#sec-variables}
+
+For our analysis, we have to use all three dataset values and thus, we chose to use data for the years 2010 to 2021, excluding 2018 and 2020, given the data was only available in all given datasets for these years.
+However our model does predict the behaviour in 2018 and 2020, using the other years’ data.
+We also created numerous csv files that isolated the specific data.
+@sec-appendix thoroughly explains the structure of the csv files.
 
@@
 #| message: false
+
 #Plot spending categories for the years
@@
 ```
-From the spending dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
+From the household spending dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
 This choice was made because these are principal expenditures for everyone.
 Recreation was added to this mix, to identify how consumers behave when inflation is high.
+@fig-total_ex_cpi indicates this relationship.
+In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
+Note for simplicity, we did not include the other categories used for total expenditures.
+For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
+However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
 
@@
 Where:
+
 * $y_i$ as the total expenditure in Canadian dollars in year i. 
@@
 
-# Appendix {-}
+# Appendix {#sec-appendix}
 

### 2024-08-08 scripts/05-graphs.R

@@ -15,17 +15,13 @@ library(scales)
 #### Load data ####
 spending <- read.csv("data/analysis_data/spending.csv")
 
-summary(spending)
 # Calculate summary statistics
 spending_stats <- spending %>%
   group_by(expenditure_categories) %>%
   summarise(
-    count = n(),
     mean_value = mean(VALUE, na.rm = TRUE),
     median_value = median(VALUE, na.rm = TRUE),
     sd_value = sd(VALUE, na.rm = TRUE),
-    min_value = min(VALUE, na.rm = TRUE),
-    max_value = max(VALUE, na.rm = TRUE),
     .groups = 'drop'
   )
 
@@ -39,12 +35,9 @@ cpi <- read.csv("data/analysis_data/cpi.csv")
 cpi_stats <- cpi %>%
   group_by(product_groups) %>%
   summarise(
-    count = n(),
     mean_value = mean(VALUE, na.rm = TRUE),
     median_value = median(VALUE, na.rm = TRUE),
     sd_value = sd(VALUE, na.rm = TRUE),
-    min_value = min(VALUE, na.rm = TRUE),
-    max_value = max(VALUE, na.rm = TRUE),
     .groups = 'drop'
   )
 

### 2024-08-08 paper/paper.qmd
Update more graphs for the data section

@@
 library(patchwork)
+
 spending <-  read_csv(here::here("data/analysis_data/spending.csv"))
+
 cpi <-  read_csv(here::here("data/analysis_data/cpi.csv"))
+
 income <- read_csv(here::here("data/analysis_data/income.csv"))
@@
 
+shelter_ex_cpi <- read_csv(here::here("data/analysis_data/shelter_ex_cpi.csv"))
+
+food_ex_cpi <- read_csv(here::here("data/analysis_data/food_ex_cpi.csv"))
+
+recreation_ex_cpi <- read_csv(here::here("data/analysis_data/recreation_ex_cpi.csv"))
+
+spending_n_income <- read_csv(here::here("data/analysis_data/spending_n_income.csv"))
+
 ```
@@
 # Introduction
+
+Consumption is the biggest indicator of a person’s needs and desires.
+Our interest is to inquire into whether tracking consumption, along with income and prices of items through inflation, will tell us a better story about a household’s financial behaviour and broader economic trends.
+By assessing these variables, we can uncover how inflation influences spending habits and whether an increase or decrease in income can mitigate or exacerbate the effects of rising or falling prices.
+The relationship between these variables can provide information in regards improving economic stability and effective social/governmental assistance to support households in managing their spending in the face of inflationary pressures.
+Policymakers can use this information to employ strategies for the wellbeing of the state.
+
+Recently, inflation has become serious problem.
+It diminishes purchasing power, which is a problem at the individual level but also state level.
+On an individual level, it affects costs of living, impeding the ability to live luxuriously.
+On a state level, it can lead to increased costs for goods and services, reducing the competitiveness of a state’s exports and making it harder for businesses to thrive on a domestic and global market.
+Thus, it is crucial states manage inflation as it will strength consumer and investor confidence, leading to economic stability and growth.
+
+This study focuses specifically on the total expenditure, household income, and the inflationary changes over a period of several years.
+
+We used a simple linear regression model and a multiple linear regression for our analysis.
+For the simple linear model, we used the the change in prices over several years against the average household spending during these years.
+For the multiple regression model, we used the average household income, along with the change in the prices to predict household spending.
+Our estimand is the
+
+Based on our model and analysis, we found that
+
+This paper was designed to communicate our model and analysis in a structured manner.
+@sec-data will provide the datasets used for this analysis.
+@sec-model will explore the simple and multiple linear regression models used for this analysis, along with the justification for the use of such models.
+@sec-results will provide the results from that analysis.
+A discussion of the results and the weaknesses and biases of this type of analysis will be provided in @sec-discussion.
 
@@
 Statistics Canada provides us with this dataset that contains the average number of people in an income category.
-For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Please note as for 2020, COVID-19 government income support and benefits are included in income estimates.
+For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Note for 2020, COVID-19 government income support and benefits are included in income estimates.
 
@@
   "Food" = "purple",
-  "Shelter" = "green",
+  "Shelter" = "darkgreen",
   "Recreation" = "blue")
@@
 Recreation was added to this mix, to identify how consumers behave when inflation is high.
-@fig-total_ex_cpi indicates this relationship.
+@fig-food_n_recreation indicates this relationship.
 In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
@@
 However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
+The only variable we use for our models from this dataset is `total expenditure`.
+However, the model can be used for the other categories.
+The model is made to be easily reproducible and can be found in the github link provided on page 1.
 
@@
 
-From the CPI dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures. The variables
+From the CPI dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures.
+The data provides cpi for specific categories, as well as all-items.
+We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending.
+The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi.
+However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19.
+Our models will use the ‘all-items’ CPI to ensure consistency.
 
@@
 #| label: fig-income
-#| fig-cap: Number of Income Earners
+#| fig-cap: Number of Income Earners per Year
 #| fig-align: "center"
@@
 This indicate all tax fillers and dependants that earn income from all sources.
-This was done to keep our analysis consistent.
+This was done to keep our analysis consistent, as we could not separate who was spending.
+Our model will use the all persons with income variable as well.
+@fig-income shows a significant increase in the number of income earners.
+This could be due to a population increase, specifically an increase in tax fillers and tax dependants.
+
+## Measurement
+
+A discussion about measurement.
 
 ```{r}
-#| label: fig-total_ex_cpi
-#| fig-cap: CPI and total expenditure over the year
+#| label: fig-total_n_shelter
+#| fig-cap: All-items CPI and total expenditure over the years (on the left), shelter CPI and shelter expenditure over the years (on the right)
 #| fig-align: "center"
@@
 
-ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
+fig_total <- ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
   geom_point(size = 1) +
-  geom_line(linewidth = 0.5, aes(group = 1)) + # Connect the points with a line
+  geom_line(linewidth = 0.5, aes(group = 1)) +
   labs(
@@
   )
+
+fig_shelter <- ggplot(shelter_ex_cpi, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5,  aes(group = 1)) +
+  labs(
+    x = "Shelter CPI",
+    y = "Shelter Expenditure",
+    color = "Year"
+  ) +
+  theme_minimal() +
+  scale_color_viridis_d() +
+  scale_x_continuous(
+    limits = c(120, 160)) +
+  scale_y_continuous(
+    labels = scales::comma,
+    limits = c(16000, 25000)
+  ) +
+  theme(
+    legend.position = "none"
+  )
+
+total_shelter_cpi <- fig_total + fig_shelter +
+  plot_layout(ncol = 2)
+
+print(total_shelter_cpi)
+
 ```
 
-# Model
+```{r}
+#| label: fig-food_n_recreation
+#| fig-cap: Food CPI and food expenditure over the years (on the left), recreation CPI and recreation expenditure over the years (on the right)
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+
+fig_food <- ggplot(food_ex_cpi, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5,  aes(group = 1)) +
+  labs(
+    x = "Food CPI",
+    y = "Food Expenditure",
+    color = "Year"
+  ) +
+  theme_minimal() +
+  scale_color_viridis_d() +
+  scale_x_continuous(
+    limits = c(120, 160)) +
+  scale_y_continuous(
+    labels = scales::comma,
+    limits = c(7000, 11000)
+  )
+
+fig_recreation <- ggplot(recreation_ex_cpi, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5, aes(group = 1)) +
+  labs(
+    x = "Recreation CPI",
+    y = "Recreation Expenditure",
+    color = "Year"
+  ) +
+  theme_minimal() +
+  scale_color_viridis_d() +
+  scale_y_continuous(
+    labels = scales::comma
+  ) +
+  theme(
+    legend.position = "none"
+  )
+
+food_recreation_cpi <- fig_food + fig_recreation+
+  plot_layout(ncol = 2)
+
+print(food_recreation_cpi)
+
+```
+
+# Model {#sec-model}
 
@@
 ```
-# Results
+
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-08-08 scripts/02-data_cleaning.R

@@ -90,7 +90,7 @@ shelter_ex_cpi <- inner_join(shelter_cpi, shelter_ex, by = "Year")
 
 #Isolate spending and CPI data for the FOOD category with just year and value
 food_ex <- spending %>%
-  filter(expenditure_categories == "Food expenditures") %>%
+  filter(expenditure_categories == "Food") %>%
   select(Year, VALUE)
 
 food_cpi <- cpi %>%
@@ -106,7 +106,7 @@ recreation_ex <- spending %>%
   select(Year, VALUE)
 
 recreation_cpi <- cpi %>%
-  filter(product_groups == "Recreation, education and reading") %>%
+  filter(product_groups == "Recreation") %>%
   select(Year, VALUE)
 
 #Combine the isolated data

### 2024-08-08 scripts/05-graphs.R

@@ -55,7 +55,7 @@ summary(income$VALUE)
 spending_colours <- c(
   "Total expenditure" = "goldenrod3", 
   "Food" = "purple",
-  "Shelter" = "green",
+  "Shelter" = "darkgreen",
   "Recreation" = "blue")
 spending_n_year <- ggplot(spending, aes(x = Year, y = VALUE, color = expenditure_categories, group = expenditure_categories)) +
   geom_line() +
@@ -82,7 +82,7 @@ print(spending_n_year)
 cpi_colours <- c(
   "All-items" = "goldenrod3", 
   "Food" = "purple",
-  "Shelter" = "green",
+  "Shelter" = "darkgreen",
   "Recreation" = "blue")
 
 cpi_n_year <- ggplot(cpi, aes(x = Year, y = VALUE, color = product_groups, group = product_groups)) +
@@ -129,7 +129,7 @@ ggplot(income, aes(x = Year, y = VALUE)) +
 #Plot spending against cpi
 ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
   geom_point(size = 1) +
-  geom_line(linewidth = 0.5, aes(group = 1)) + # Connect the points with a line
+  geom_line(linewidth = 0.5, aes(group = 1)) +
   labs(
     x = "All-Items CPI",
     y = "Total Expenditure",
@@ -144,45 +144,66 @@ ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
     limits = c(75000, 100000)
   )
 
-ggplot(shelter_ex_cpi, aes(x = VALUE.x, y = VALUE.y)) +
-  geom_point() +
+ggplot(shelter_ex_cpi, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5,  aes(group = 1)) +
   labs(
-    title = "Shelter Expenditure vs Shelter CPI",
     x = "Shelter CPI",
-    y = "Shelter Expenditure"
+    y = "Shelter Expenditure",
+    color = "Year"
   ) +
-  theme_minimal()
+  theme_minimal() +
+  scale_color_viridis_d() +
+  scale_x_continuous(
+    limits = c(120, 160)) +
+  scale_y_continuous(
+    labels = scales::comma,
+    limits = c(16000, 25000)
+  )
 
[…]

### 2024-08-08 paper/paper.qmd
Rearrange data section

@@
 # Data {#sec-data}
-## Sources
+
+## Methodology
+The language and environment used for this analysis is R (R Core Team 2020), alongside the tidyverse (cite), rstanarm (cite), cansim (cite), knitr (cite), ggplot2 (cite), scales (cite), modelsummary (cite), and marginaleffects (cite) packages.
+
+We downloaded the data directly from Statistics Canada using the cansim package. We cleaned it, identified missing values and wrote csv files to make it easily accessible using R studio.
+
+## Sources and Variables {#sec-sources_variables}
 A good indicator of inflation is the Consumer Price Index (CPI), as it measures the change in price using a fixed basket of goods and services.
@@
 The expenditure and cpi data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question.
-@sec-variables will illustrate the visualised data.
+@sec-sources_variables will illustrate the visualised data.
 
@@
 
-### Inflation Data
-
-`cpi.csv`
-
-The dataset provides CPI for every year.
-It is not seasonally adjusted for changes as it is an annual average.
-Thus, it is actual price paid by consumers and not adjusted for seasonal fluctuations.
-The base year for the CPI calculation is 2002 where CPI is 100.
-Thus, if our CPI is 120 for a year, it indicates a 20% increase since 2002.
-This value is useful as an indicator for inflation, as it provides the increase and comparison between years, rather than a absolute price values.
-
-### Income Data
-
-`income.csv`
-
-Throughout the province of Ontario, there exits people in many different financial situations.
-Statistics Canada provides us with this dataset that contains the average number of people in an income category.
-For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Note for 2020, COVID-19 government income support and benefits are included in income estimates.
-
-## Methodology
-The language and environment used for this analysis is R (R Core Team 2020), alongside the tidyverse (cite), rstanarm (cite), cansim (cite), knitr (cite), ggplot2 (cite), scales (cite), modelsummary (cite), and marginaleffects (cite) packages.
-
-We downloaded the data directly from Statistics Canada using the cansim package. We cleaned it, identified missing values and wrote csv files to make it easily accessible using R studio.
-
-## Variables {#sec-variables}
-
-For our analysis, we have to use all three dataset values and thus, we chose to use data for the years 2010 to 2021, excluding 2018 and 2020, given the data was only available in all given datasets for these years.
-However our model does predict the behaviour in 2018 and 2020, using the other years’ data.
-We also created numerous csv files that isolated the specific data.
-@sec-appendix thoroughly explains the structure of the csv files.
+From this dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
+This choice was made because these are principal expenditures for everyone.
+Recreation was added to this mix, to identify how consumers behave when inflation is high.
+@fig-food_n_recreation indicates this relationship.
+In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
+Note for simplicity, we did not include the other categories used for total expenditures.
+For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
+However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
+The only variable we use for our models from this dataset is `total expenditure`.
+However, the model can be used for the other categories.
+The model is made to be easily reproducible and can be found in the github link provided on page 1.
 
@@
 ```
-From the household spending dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
-This choice was made because these are principal expenditures for everyone.
-Recreation was added to this mix, to identify how consumers behave when inflation is high.
-@fig-food_n_recreation indicates this relationship.
-In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
-Note for simplicity, we did not include the other categories used for total expenditures.
-For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
-However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
-The only variable we use for our models from this dataset is `total expenditure`.
-However, the model can be used for the other categories.
-The model is made to be easily reproducible and can be found in the github link provided on page 1.
+
+### Inflation Data
+
+`cpi.csv`
+
+The dataset provides CPI for every year.
+It is not seasonally adjusted for changes as it is an annual average.
+Thus, it is actual price paid by consumers and not adjusted for seasonal fluctuations.
+The base year for the CPI calculation is 2002 where CPI is 100.
+Thus, if our CPI is 120 for a year, it indicates a 20% increase since 2002.
+This value is useful as an indicator for inflation, as it provides the increase and comparison between years, rather than a absolute price values.
+
+From this dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures.
+The data provides cpi for specific categories, as well as all-items.
+We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending.
+The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi.
+However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19.
+Our models will use the ‘all-items’ CPI to ensure consistency.
 
@@
 
-From the CPI dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures.
-The data provides cpi for specific categories, as well as all-items.
-We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending.
-The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi.
-However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19.
-Our models will use the ‘all-items’ CPI to ensure consistency.
+### Income Data
+
+`income.csv`
+
+Throughout the province of Ontario, there exits people in many different financial situations.
+Statistics Canada provides us with this dataset that contains the average number of people in an income category.
+For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Note for 2020, COVID-19 government income support and benefits are included in income estimates.
+
+From this dataset, we chose to isolate the values for the category “all persons with income”.
+This indicate all tax fillers and dependants that earn income from all sources.
+This was done to keep our analysis consistent, as we could not separate who was spending.
+Our model will use the all persons with income variable as well.
+@fig-income shows a significant increase in the number of income earners.
+This could be due to a population increase, specifically an increase in tax fillers and tax dependants.
 
@@
 
-From the income dataset, we chose to isolate the values for the category “all persons with income”.
-This indicate all tax fillers and dependants that earn income from all sources.
-This was done to keep our analysis consistent, as we could not separate who was spending.
-Our model will use the all persons with income variable as well.
-@fig-income shows a significant increase in the number of income earners.
-This could be due to a population increase, specifically an increase in tax fillers and tax dependants.
+For our analysis, we have to use all three dataset values and thus, we chose to use data for the years 2010 to 2021, excluding 2018 and 2020, given the data was only available in all given datasets for these years.
+However our model does predict the behaviour in 2018 and 2020, using the other years’ data.
+We also created numerous csv files that isolated the specific data.
+@sec-appendix thoroughly explains the structure of the csv files.
 

### 2024-08-09 paper/paper.qmd
Update model

@@
 ## Methodology
-The language and environment used for this analysis is R (R Core Team 2020), alongside the tidyverse (cite), rstanarm (cite), cansim (cite), knitr (cite), ggplot2 (cite), scales (cite), modelsummary (cite), and marginaleffects (cite) packages.
+The language and environment used for this analysis is R [@citeR], alongside the tidyverse (@tidyverse), rstanarm (@rstanarm), cansim [@cansim], ggplot2 [@ggplot2], scales [@scales], modelsummary [@modelsummary], and marginaleffects [@marginaleffects] packages.
 
@@
 
-* $y_i$ as the total expenditure in Canadian dollars in year i. 
-* $cpi_i$ is the annual Consumer Price Index (CPI) for year i
+* $y_i$ is the continuous outcome variable, representing the total expenditure in Canadian dollars in year i. 
+* $cpi_i$ is the predictor variable, representing the annual Consumer Price Index (CPI) for year i
 
 ### Multiple linear regression
-The model we employ is:
+The model we employed is:
 
@@
 
-* $y_i$ as the total expenditure in Canadian dollars in year i. 
-* $cpi_i$ is the annual Consumer Price Index (CPI) for year i
-* $income_i$ is the number of people in Ontario earning an income
+* $y_i$ is the continuous outcome variable, representing the total expenditure in Canadian dollars in year $i$. 
+* $cpi_i$ is the predictor variable, representing the annual Consumer Price Index (CPI) for year $i$
+* $income_i$ is the number of people earning an income in year $i$.
 

### 2024-08-10 scripts/00-simulate_data.R
Simulate and test data

@@ -7,13 +7,32 @@
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]
 
-
 #### Workspace setup ####
+
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
 
+set.seed(5678)
+
+years <- seq(1990, 2021, by = 1)
+
+expenditure_predictions <- tibble(
+  Year = years,
+  cpi = rnorm(length(years), mean = 120, sd = 10),
+  income = round(rnorm(length(years), mean = 10000, sd = 1500))
+)
+
+print(expenditure_predictions)
+
+#### Test simulated data ####
+
+nrow(expenditure_predictions) == 32
+
+expenditure_predictions$Year |> class() == "numeric"
+
+expenditure_predictions$cpi |> class() == "numeric"
+
+expenditure_predictions$income |> class() == "integer"
 
 

### 2024-08-11 paper/paper.qmd
Update model to complete prior and posterior checks

@@
 library(patchwork)
+library(rstanarm)
 
@@
 In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
-Note for simplicity, we did not include the other categories used for total expenditures.
+Note for simplicity, we did not include the other categories used to calculate total expenditures in the original dataset.
+This is the reason the total expenditure is significantly higher than the other categories.
 For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
@@
 A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
+A discussion about measurement.
 
@@
 
-The goal of our model is to forecasting the total expenditure in a household in a year, based solely on a the Consumer Price Index, and the number of people in Ontario earning an income.
+In this section we build a model to investigate household consumer behaviour.
+
+## Model set-up
+
+The goal of our model is to forecast the total expenditure in a household in a year, based solely on the Consumer Price Index of all-items, and the number of people in Ontario earning an income.
 We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
 We use the default priors from `rstanarm`.
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
+Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 
-### Model justification
+## Model justification
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
+From @fig-total_n_shelter and fig-food_n_recreation, we see a positive relationship in expenditure and CPI as the years pass.
+Implying that despite rising prices, household spending has continued to increase, specifically in the food and shelter categories.
+Due to this relationship, we anticipate the relationship is close to linear.
+We assume this relationship because food and shelter are necessities unlike recreational spending.
 
@@
 #| label: fig-2
-#| fig-cap: Linear model
+#| fig-cap: Relationship between CPI for all-items and total household expenditure
 #| echo: false
@@
   labs(
-    title = "Total Expenditure vs All-Items CPI",
     x = "All-Items CPI",
@@
 #| label: fig-3
-#| fig-cap: Linear model
+#| fig-cap: Relationship between CPI for all-items and total household expenditure
 #| echo: false
@@
 
-total_ex_cpi |>
-  ggplot(aes(x = cpi, y = expenditure, color = factor(Year))) +
-  geom_point(size = 1) +
-  geom_smooth(
-    method = "lm",
-    se = TRUE,
-    color = "black",
-    linetype = "dashed",
-    formula = y ~ x
-  ) +
-  labs(
-    title = "Total Expenditure vs All-Items CPI",
-    x = "All-Items CPI",
-    y = "Total Expenditure",
-    color = "Year"
-  ) +
-  scale_y_continuous(labels = scales::comma) +
-  scale_color_viridis_d() +
-  theme_minimal()
 ```
@@
 
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-second_model <-
-  readRDS(file = here::here("models/second_model.rds"))
-```
+Our model results are summarized in @tbl-modelresults.
+The table compares our two models, each predicting the outcome variable (total expenditure), based on either CPI only, and CPI and the number of income earners.
 
@@
 
+first_model <-
+  readRDS(file = here::here("models/first_model.rds"))
+second_model <-
+  readRDS(file = here::here("models/second_model.rds"))
+
 modelsummary::modelsummary(
   list(
-    "First model" = first_model,
-    "Second model" = second_model
+    "CPI" = first_model,
+    "Add Income Earners" = second_model
   ),

### 2024-08-11 scripts/03-test_data.R

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
+# Author: Rahma Binth Mohammad
 # Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Run the "00-install_packages.R", "01-download_data", and "02-data_cleaning"
 # Any other information needed? [...UPDATE THIS...]

### 2024-08-11 scripts/04-model.R

@@ -24,7 +24,7 @@ first_model <-
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+    seed = 5678
   )
 
 second_model <-
@@ -35,7 +35,7 @@ second_model <-
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+    seed = 5678
   )
   
 #### Save model ####

### 2024-08-12 scripts/00-install_packages.R
Rename files

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Installs the necessary packages. 
 # Author: Rahma Binth Mohammad
-# Date: 
+# Date: July 20, 2024
 # Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
 
@@ -9,4 +9,4 @@
 install.packages("tidyverse")
 install.packages("cansim")
 install.packages("dplyr")
-
+install.packages("rstanarm")

### 2024-08-12 scripts/01-simulate_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates the dataset
+# Author: Rahma Binth Mohammad
+# Date: July 20, 2024
+# Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Run "00-install_packages.R."
 
 #### Workspace setup ####
 

### 2024-08-12 scripts/02-download_data.R

@@ -1,17 +1,16 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Statistics Canada
 # Author: Rahma Binth Mohammad
-# Date: 11 February 2023 [...UPDATE THIS...]
+# Date: July 20, 2024
 # Contact: rahma.binthmohammad@mail.utoronto.ca
 # Pre-requisites: Run "00-install_packages.R".
-# Any other information needed? [...UPDATE THIS...]
 
 #### Workspace setup ####
 library(cansim)
 library(tidyverse)
 library(dplyr)
 
-#### Download data ####
+#### Download data from Statistics Canada ####
 
 #Household spending data
 spending <- get_cansim("11-10-0222-01")
@@ -19,7 +18,8 @@ spending <- get_cansim("11-10-0222-01")
 #Annual CPI data
 cpi <- get_cansim("18-10-0005-01")
 
-#Household income, Data is too large to push to git, hence it is downloaded. You can use this code to get it directly.
+#Household income, Data is too large to push to git, hence it is downloaded. 
+#You can use this code to get it directly. 
 #income <- get_cansim("11-10-0008-01")
 
 #### Save data ###

### 2024-08-12 scripts/03-data_cleaning.R

@@ -1,18 +1,18 @@
 #### Preamble ####
 # Purpose: Cleans the raw plane data recorded by Statistics Canada and filters the necessary columns
 # Author: Rahma Binth Mohammad
-# Date: 6 April 2023 [...UPDATE THIS...]
+# Date: Jul 20, 2024
 # Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Run "00-install_packages.R" and "01-download_data". 
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Run "00-install_packages.R" and "02-download_data.R."
 
 #### Workspace setup ####
+
 library(tidyverse)
 
 #### Clean data ####
+#Read, filter and rename the household spending data
 
-#Filter and rename the household spending data
 spending <- read.csv("data/raw_data/spending.csv")
 
 spending <- spending %>% 
@@ -29,7 +29,7 @@ spending <- spending %>%
     "Food expenditures" = "Food"
   ))
 
-#Filter and rename the annual CPI data
+#Read, filter and rename the annual CPI data
 cpi <- read.csv("data/raw_data/cpi.csv")
 
 cpi <- cpi %>% 
@@ -46,9 +46,7 @@ cpi <- cpi %>%
     "Recreation, education and reading" = "Recreation"
   ))
 
-#Filter and rename the household income data
-income <- read.csv("data/raw_data/income.csv")
-
+#Read, filter and rename the household income data
 income <- income %>% 
   rename(income_categories = Persons.with.income, Year = REF_DATE, age = Age.group)
 

### 2024-08-12 scripts/05-model.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
+# Purpose: Models the data two different ways.
 # Author: Rahma Binth Mohammad
-# Date: 11 February 2023 [...UPDATE THIS...]
+# Date: July 24, 2024
 # Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Run the "00-install_packages.R", "01-download_data.R", "02-data_cleaning.R", "03-test_data.R"
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Run the "00-install_packages.R", "02-download_data.R", and "03-data_cleaning.R" files. 
 
 #### Workspace setup ####
 library(tidyverse)
@@ -48,4 +47,3 @@ saveRDS(
   second_model,
   file = "models/second_model.rds"
 )
-

### 2024-08-12 scripts/06-graphs.R

@@ -1,18 +1,17 @@
 #### Preamble ####
 # Purpose: Creates graphs with filtered data that is included in the paper, useful for reproduction
 # Author: Rahma Binth Mohammad
-# Date: 11 February 2023 [...UPDATE THIS...]
+# Date: July 24, 2024
 # Contact: rahma.binthmohammad@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Run the "00-install_packages.R", "02-download_data.R", and "03-data_cleaning.R" files. 
 
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
 library(scales)
 
-#### Load data ####
+#### Read data ####
 spending <- read.csv("data/analysis_data/spending.csv")
 
 # Calculate summary statistics
@@ -77,6 +76,7 @@ spending_n_year <- ggplot(spending, aes(x = Year, y = VALUE, color = expenditure
     )
 
 print(spending_n_year)
+
 #Plot CPI values for each category for each year
 #Set colours so that it is easily comparible with spending data
 cpi_colours <- c(
@@ -126,7 +126,7 @@ ggplot(income, aes(x = Year, y = VALUE)) +
     limits = c(9600000, 11250000)
   )
 
-#Plot spending against cpi
+#Plot all spending categories against all cpi categories
 ggplot(total_ex_cpi, aes(x = cpi, y = expenditure, color = factor(Year))) +
   geom_point(size = 1) +
   geom_line(linewidth = 0.5, aes(group = 1)) +
@@ -192,10 +192,10 @@ ggplot(recreation_ex_cpi, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
     labels = scales::comma
   )
 
-# Plot with points and lines
+# Plot with spending against income
 ggplot(spending_n_income, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
   geom_point(size = 1) +
-  geom_line(linewidth = 0.5, aes(group = 1)) + # Connect the points with a line
+  geom_line(linewidth = 0.5, aes(group = 1)) + 
   labs(
     x = "Number of Income Earners",
     y = "Total Expenditure",

### 2024-08-12 paper/paper.qmd
Update LLM Usage

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Does inflation make you spend less?"
 author: 
@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 
+#Load packages
 library(tidyverse)
@@
 
+# Read in files from data/analysis_data
 spending <-  read_csv(here::here("data/analysis_data/spending.csv"))
+cpi <-  read_csv(here::here("data/analysis_data/cpi.csv"))
+income <- read_csv(here::here("data/analysis_data/income.csv"))
+total_ex_cpi <- read_csv(here::here("data/analysis_data/total_ex_cpi.csv")) 
+shelter_ex_cpi <- read_csv(here::here("data/analysis_data/shelter_ex_cpi.csv"))
+food_ex_cpi <- read_csv(here::here("data/analysis_data/food_ex_cpi.csv"))
+recreation_ex_cpi <- read_csv(here::here("data/analysis_data/recreation_ex_cpi.csv"))
+spending_n_income <- read_csv(here::here("data/analysis_data/spending_n_income.csv"))
+```
+# Introduction
+Consumption is the biggest indicator of a person’s needs and desires.
+It tells us the exact amount and item people are willing to spend on.
+Our interest is to inquire into whether tracking consumption, along with income and prices of items through inflation, will tell us a better story about a household’s financial behaviour and broader economic trends.
+By assessing these variables, we can uncover how inflation influences spending habits and whether an increase or decrease in the number of income earners in a region can mitigate or exacerbate the effects of rising or falling prices.
+The relationship between these variables can provide information in regards to improving economic stability and effective social governmental assistance to support households in managing their spending in the face of inflationary pressures.
+Overall, policymakers can use this information to employ strategies for the wellbeing of the state and ensure households and even states don’t go deeper into debt.
 
-cpi <-  read_csv(here::here("data/analysis_data/cpi.csv"))
-
-income <- read_csv(here::here("data/analysis_data/income.csv"))
-
-total_ex_cpi <- read_csv(here::here("data/analysis_data/total_ex_cpi.csv"))
-
-shelter_ex_cpi <- read_csv(here::here("data/analysis_data/shelter_ex_cpi.csv"))
-
-food_ex_cpi <- read_csv(here::here("data/analysis_data/food_ex_cpi.csv"))
-
-recreation_ex_cpi <- read_csv(here::here("data/analysis_data/recreation_ex_cpi.csv"))
-
-spending_n_income <- read_csv(here::here("data/analysis_data/spending_n_income.csv"))
-
-```
-
-# Introduction
-
-Consumption is the biggest indicator of a person’s needs and desires.
-Our interest is to inquire into whether tracking consumption, along with income and prices of items through inflation, will tell us a better story about a household’s financial behaviour and broader economic trends.
-By assessing these variables, we can uncover how inflation influences spending habits and whether an increase or decrease in income can mitigate or exacerbate the effects of rising or falling prices.
-The relationship between these variables can provide information in regards improving economic stability and effective social/governmental assistance to support households in managing their spending in the face of inflationary pressures.
-Policymakers can use this information to employ strategies for the wellbeing of the state.
-
-Recently, inflation has become serious problem.
+Recently, inflation has become serious issue.
 It diminishes purchasing power, which is a problem at the individual level but also state level.
@@
 
-This study focuses specifically on the total expenditure, household income, and the inflationary changes over a period of several years.
-
 We used a simple linear regression model and a multiple linear regression for our analysis.
 For the simple linear model, we used the the change in prices over several years against the average household spending during these years.
-For the multiple regression model, we used the average household income, along with the change in the prices to predict household spending.
-Our estimand is the
+For the multiple regression model, we used the number of income earners, along with the change in the prices to predict household spending.
+Our estimand is the average total household expenditure including recreational usage.
 
-Based on our model and analysis, we found that
+Based on our model and analysis, we found that despite rising prices, households continued to spend money on necessary items, as well as recreational spending.
+However, recreational spending saw a slower increase over a decade.
+Incorporating the change in the number of income earners ensured we factored in the change in population and the constant flux of people moving to or away from the region.
 
@@
 # Data {#sec-data}
+## Methodology
+The language and environment used for this analysis is R [@citeR], alongside the tidyverse (@tidyverse), rstanarm (@rstanarm), cansim [@cansim], ggplot2 [@ggplot2], scales [@scales], and modelsummary [@modelsummary] packages.
 
-## Methodology
-The language and environment used for this analysis is R [@citeR], alongside the tidyverse (@tidyverse), rstanarm (@rstanarm), cansim [@cansim], ggplot2 [@ggplot2], scales [@scales], modelsummary [@modelsummary], and marginaleffects [@marginaleffects] packages.
+We downloaded three different datasets directly from Statistics Canada using the cansim package.
+We cleaned it by isolating the variables we need, removed missing values, and wrote csv files to make it easily accessible using R studio.
 
-We downloaded the data directly from Statistics Canada using the cansim package. We cleaned it, identified missing values and wrote csv files to make it easily accessible using R studio.
-
-## Sources and Variables {#sec-sources_variables}
+## Sources {#sec-sources}
 A good indicator of inflation is the Consumer Price Index (CPI), as it measures the change in price using a fixed basket of goods and services.
-Statistics Canada is a government agency that collects information on Canada’s economy (cite).
-Thus, it has the data that provides the CPI for numerous years.
-Statistics Canada also provides us with other datasets such as the average household income and the average household expenditures, in a year for all the provinces.
-We have chosen to look at Ontario solely, as we hold firmly Ontario behaves very differently due to its high population in relation to the other provinces.
+Statistics Canada is a government agency that collects information on Canada’s economy [@statcan] and it has the data that provides the CPI for numerous years.
+Statistics Canada also provides us with other datasets such as the average household expenditures in a year for all the provinces and the number of income earners in a province.
+We have chosen to look at Ontario solely, as we hold firmly that Ontario behaves very differently due to its high population in relation to the other provinces.
 The data for Canada as a whole is complicated as there are many missing values for some provinces.
-Our analysis and the data provided by Statistics Canada makes it possible to reproduce our model for different provinces using the github repository.
-While there are other datasets available, these three datasets are provided by the same source and has data for provinces across Canada.
-The expenditure and cpi data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question.
-@sec-sources_variables will illustrate the visualised data.
+However, our analysis and the data provided by Statistics Canada makes it possible to reproduce our model for different provinces using the github repository for this paper.
+While there are other datasets available, these three datasets are provided by the same source and has data for provinces across Canada, making it easily reproducible.
+The expenditure and CPI data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question.
+@sec-sources will illustrate the visualised data.
 
 This paper’s objective is to address the following questions using data collected from a government census: (1) Does changes in prices affect expenditure?
-(2) Does changed in prices affect expenditure as income changes?
+(2) Does changed in prices affect expenditure as the number of income earners change?
 
+## Variables and Measurement
 ### Spending Data
-
-`spending.csv`
-
-This dataset contains the average total expenditures of the Ontario resident for the years 2010 to 2021, excluding years 2018 and 2020.
-It also contains the breakdown of the expenditures, food purchased (from restaurant and stores), shelter (rented and owned), utilities such as water, fuel, and electricity for the principal accommodation, clothing, transportation (private and public), healthcare, recreational goods and services.
-
-From this dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures.
-This choice was made because these are principal expenditures for everyone.
-Recreation was added to this mix, to identify how consumers behave when inflation is high.
-@fig-food_n_recreation indicates this relationship.
-In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
-Note for simplicity, we did not include the other categories used to calculate total expenditures in the original dataset.
-This is the reason the total expenditure is significantly higher than the other categories.
-For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
-However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
-The only variable we use for our models from this dataset is `total expenditure`.
-However, the model can be used for the other categories.
-The model is made to be easily reproducible and can be found in the github link provided on page 1.
 
@@
 
+`spending.csv`
+
+This dataset contains the average total expenditures of the Ontario resident for the years 2010 to 2021, excluding years 2018 and 2020.
+It also contains the breakdown of the expenditures, food purchased (from restaurant and stores), shelter (rented and owned), utilities such as water, fuel, and electricity for the principal accommodation, clothing, transportation (private and public), healthcare, recreational goods and services.
+
+From this dataset, we chose to select the values for all food expenditures, all shelter expenditures, all recreation expenditures, and total expenditures.
+This choice was made because these are principal expenditures for everyone.
+Recreation was added to this mix, to identify how consumers behave when inflation is high.
+It helped answer the question of whether consumers spend less on items that are unnecessary to survive?
+We define necessary items to be food, shelter, and clothing.
+
+In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth.
+Note for simplicity, we did not include the other categories used to calculate total expenditures in the original dataset.
+This is the reason the total expenditure is significantly higher than the other categories.
+For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers.
+However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years.
+The only variable we use for our models from this dataset is `total expenditure`.
+However, the model can be used for the other categories.
+The model is made to be easily reproducible and can be found in the github link provided.
+
 ### Inflation Data
-
-`cpi.csv`
-
-The dataset provides CPI for every year.
-It is not seasonally adjusted for changes as it is an annual average.
-Thus, it is actual price paid by consumers and not adjusted for seasonal fluctuations.
-The base year for the CPI calculation is 2002 where CPI is 100.
-Thus, if our CPI is 120 for a year, it indicates a 20% increase since 2002.
-This value is useful as an indicator for inflation, as it provides the increase and comparison between years, rather than a absolute price values.
-
-From this dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures.
-The data provides cpi for specific categories, as well as all-items.
-We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending.
-The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi.
-However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19.
-Our models will use the ‘all-items’ CPI to ensure consistency.
 
@@
 
+`cpi.csv`
+
+The dataset provides CPI for every year.
+It is not seasonally adjusted for changes as it is an annual average.
+Thus, it is the actual price paid by consumers and not adjusted for seasonal fluctuations.
+The base year for the CPI calculation is 2002 where CPI is 100.
+Thus, if our CPI is 120 for a year, it indicates a 20% increase since 2002.
+This value is useful as an indicator for inflation, as it provides the increase and comparison between years, rather than a absolute price values.
+
+We chose to look at the annual average for CPI to keep it consistent with the annual average of expenditures.
+The data provides CPI for specific categories, as well as all-items.
+We chose to select CPI for all-items, food, shelter, and recreations, for comparison with the household spending.
+The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi.
+However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19.
+Our models will use the ‘all-items’ CPI to ensure consistency.
+
 ### Income Data
-
-`income.csv`
-
-Throughout the province of Ontario, there exits people in many different financial situations.
-Statistics Canada provides us with this dataset that contains the average number of people in an income category.
-For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Note for 2020, COVID-19 government income support and benefits are included in income estimates.
-
-From this dataset, we chose to isolate the values for the category “all persons with income”.
-This indicate all tax fillers and dependants that earn income from all sources.
-This was done to keep our analysis consistent, as we could not separate who was spending.
-Our model will use the all persons with income variable as well.
-@fig-income shows a significant increase in the number of income earners.
-This could be due to a population increase, specifically an increase in tax fillers and tax dependants.
 
@@
 ```
+`income.csv`
 
-For our analysis, we have to use all three dataset values and thus, we chose to use data for the years 2010 to 2021, excluding 2018 and 2020, given the data was only available in all given datasets for these years.
+Everyone’s financial situation is different, especially in Ontario.
+Statistics Canada provides us with this dataset that contains the average number of people in an income category.
+For example, “persons with income under $5000", “persons with income of $5000 and over”, “persons with income of $10 000 and over.” Note for 2020, COVID-19 government income support and benefits are included in income estimates.
+
+From this dataset, we chose to isolate the values for the category “all persons with income”.
+This indicate all tax fillers and dependants that earn income from all sources.
+This was done to keep our analysis consistent, as we could not separate who was spending on what item.
+Our model will use the `all persons with income` variable as well.
+@fig-income shows a significant increase in the number of income earners.
+This could be due to a population increase, specifically an increase in tax fillers and tax dependants because the data is taken directly from filed tax returns.
+
+For our analysis, we had to use all three dataset values and thus, we chose to use data for the years 2010 to 2021, excluding 2018 and 2020, given the data was only available in all given datasets for these years.
 However our model does predict the behaviour in 2018 and 2020, using the other years’ data.
 We also created numerous csv files that isolated the specific data.
-@sec-appendix thoroughly explains the structure of the csv files.
-
-## Measurement
-
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
-A discussion about measurement.
+The variables discussed in this section became an entry in our dataset, because each variable answers a relevant question.
 
@@
 ```
+```{r}
+#| label: fig-spending_n_income
+#| fig-cap: The number of income earners and their average expenditures
+#| fig-align: "center"
+#| echo: false
+#| warning: false
+#| message: false
+ggplot(spending_n_income, aes(x = VALUE.x, y = VALUE.y, color = factor(Year))) +
+  geom_point(size = 1) +
+  geom_line(linewidth = 0.5, aes(group = 1)) + 
+  labs(
+    x = "Number of Income Earners",
+    y = "Total Expenditure",
+    color = "Year") +
+  theme_minimal() +
+  scale_color_viridis_d() +
+  scale_x_continuous(
+    labels = scales::comma) +
+  scale_y_continuous(
+    labels = scales::comma)
+```
+
+The variables discussed in @sec-sources became an entry in our dataset, because each variable answers a relevant question.
+Household spending data was only available from years 2010-2021.
+This is a relatively small interval so we had limitations for our model.
+However, it is still useful because a lot can change in a decade and if you look at data before 2010, it will not give you a clear visual of what is happening today.
+Thus, making it harder to make political or economical decisions that impact today’s world.
+
+@fig-total_n_shelter, @fig-food_n_recreation, and @fig-spending_n_income are all the predictor variables graphed against the total expenditure.
+There is an increase in spending over the years, as well as CPI and income earners.
+In the rare occasions, where CPI decreases, expenditure decreases as well.
 
@@
 
-In this section we build a model to investigate household consumer behaviour.
-
-## Model set-up
-
-The goal of our model is to forecast the total expenditure in a household in a year, based solely on the Consumer Price Index of all-items, and the number of people in Ontario earning an income.
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
-We use the default priors from `rstanarm`.
-Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-### Simple linear regression
-The model we employ is:
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \times \mbox{cpi}_i + \epsilon\\
-\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
-\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-Where:
-
-* $y_i$ is the continuous outcome variable, representing the total expenditure in Canadian dollars in year i. 
-* $cpi_i$ is the predictor variable, representing the annual Consumer Price Index (CPI) for year i
-
-### Multiple linear regression
-The model we employed is:
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 \times \mbox{cpi}_i + \beta_2 \mbox{income}_i+ \epsilon\\
-\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
-\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
-\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-Where:
-
-* $y_i$ is the continuous outcome variable, representing the total expenditure in Canadian dollars in year $i$. 
-* $cpi_i$ is the predictor variable, representing the annual Consumer Price Index (CPI) for year $i$
-* $income_i$ is the number of people earning an income in year $i$.
-
-## Model justification
-
-From @fig-total_n_shelter and fig-food_n_recreation, we see a positive relationship in expenditure and CPI as the years pass.
-Implying that despite rising prices, household spending has continued to increase, specifically in the food and shelter categories.
-Due to this relationship, we anticipate the relationship is close to linear.
-We assume this relationship because food and shelter are necessities unlike recreational spending.
+In this section, we build a model to investigate household consumer behaviour.
 
@@
 
-```{r}
-#| label: fig-3
-#| fig-cap: Relationship between CPI for all-items and total household expenditure
-#| echo: false
-#| warning: false
-#| message: false
-library(scales)
+## Model set-up
 
-```
+The goal of our model is to forecast the total expenditure in a household in a year, based solely on the Consumer Price Index of all-items, and the number of people in Ontario earning an income.
+We run the model in R [@citeR] using the `rstanarm` package of @rstanarm.
+
+### Simple linear regression
+The model we employ is:
+
+\begin{align} 
+y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 + \beta_1 \mbox{cpi}_i + \epsilon\\
+\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1)
+\end{align}
+
+Where:
+
+* $y_i$ is the continuous outcome variable, representing the total expenditure in Canadian dollars in year i. 
+* $cpi_i$ is the predictor variable, representing the annual Consumer Price Index (CPI) for year i
+
+### Multiple linear regression
+In our second model, we added another predictor variable to employ a multiple linear regression model.
+
+\begin{align} 
+y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 + \beta_1 \mbox{cpi}_i + \beta_2 \mbox{income}_i+ \epsilon\\
+\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1)
+\end{align}
+
+Where:
+
+* $income_i$ is the another predictor variable, representing the number of people earning an income in year $i$.
+
+## Model justification
+
+From @fig-total_n_shelter and @fig-food_n_recreation, we see a positive relationship in expenditure and CPI as the years pass.
+Implying that despite rising prices, household spending has continued to increase, specifically in the food and shelter categories.
+Due to this relationship, we anticipate the relationship is close to linear.
+We assume this relationship because food and shelter are necessities unlike recreational spending.
 
 # Results {#sec-results}
-
-Our model results are summarized in @tbl-modelresults.
-The table compares our two models, each predicting the outcome variable (total expenditure), based on either CPI only, and CPI and the number of income earners.
 
@@
 #| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of "
+#| tbl-cap: "Explanatory models of expenditure "
 #| warning: false
@@
 
-```{r}
+Since linear models tend to focus on either inference or prediction [@rohan], we have used it infer that expenditure will continue to increase as prices increase.
+Our model results are summarized in @tbl-modelresults.
+The table compares our two models, each predicting the outcome variable (total expenditure), based on either CPI only, or CPI and the number of income earners.
 
-```
+In @tbl-modelresults, we see the coefficient for the CPI is approximately 927, implying that for every unit increase in CPI, the total expenditure increases by approximately 927 units.
+In simple terms, CPI is a measure of inflation, and as inflation increases, expenditure increases.
+However, for the second model, we see a significantly smaller increase of 227 units in total expenditure.
+This illustrates that with the addition of the second predictor variable (number of income earners), we were able to get a more accurate representation because expenditure naturally increases as the number of income earners increases, due to the assumption of lifestyle inflation, people will spend more money if they make more money.
+@tbl-modelresults also indicates the coefficient for income variable, which is quiet small, 0.01.
+However, this number did not worry us, as the scale is important in this context.
+The number of income earners is in the millions while the total expenditure is in the thousands.
+Since the coefficient is positive, we can predict that as the number of the income earners increase, the total expenditure would increase as well.
+
+The first model provides simplicity and ease, while the second model includes another factor for review.
+The first model is the ideal fit as the standard errors are smaller.
+Policymakers can also make decisions solely on the impact of CPI alone.
+The number of income earners is relevant.
+However, from the coefficient for the income variable, we can conclude the average household income is not impacted too much.
+Expenditure is a result of many factors, household size, seasons, income range of the household, and the population and number of income earners definitely plays a role in how expenditure can increase.
 
 # Discussion {#sec-discussion}
+## Findings
+The data and variables were discussed in @sec-data and the @sec-results identified the results of our model in @sec-model.
+Using all this, we identified our main findings.
+A common theory is that as inflation rises, people spend less.
+We conclude this theory is false because people have to spend more to survive, as found from our model results.
+We saw some decreases in recreation but overall, it increased.
+This could be do to the fact that technology and social media, has been widely used to make more money in less time, unlike a 9am to 5pm job.
+Thus, as prices increase, people must be making more money or borrowing money to make due an spend on necessities like food and shelter.
 
-## First discussion point {#sec-first-point}
+## Bias
+There are also other factors to consider that influence total expenditure.
+For example, unemployment rate and interest rates.
+For simplicity and due to data availability, we chose it to include two predictor variables only, since prices have a significant impact on spending.
+For the spending and CPI data, we used aggregate data to keep it consistent and broad.
+However, this can mask the variations in each dataset.
+For example, spending and CPI on entertainment and transportation look different than food and shelter.
+Thus, there can be misleading averages in a year.
+From the income data, we chose “all persons with income” which can undermine the experiences of marginalized vulnerable groups in the region.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+## Weaknesses
+This paper focused explicitly on the total expenditure, inflationary changes over a period of several years, and the number of income earners in Ontario.
+Household spending data was only available from years 2010-2021.
+This is a relatively small interval so we had limitations for our model.
+However, it is still useful because a lot can change in a decade and if you look at data before 2010, it will not give you a clear visual of what is happening today.
+Thus, making it harder to make political or economical decisions that impact today’s world.
 
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+## Policy Implications and Future Research 
+It is important to find out whether people are earning more income or depleting their savings, or borrowing more than they can handle.
+With this information, policymakers can identify a clear problem, and that is the first step in developing a solution for the problem.
+If not, the region can face a significant economical collapse.
 
@@
 
-\appendix
-
-# Appendix {#sec-appendix}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
+# References

### 2024-08-12 paper/paper.qmd
Add abstract and fix grammatical errors

@@
 thanks: "Code and data are available at: https://github.com/rahmabinth/consumer-behaviour"
-date: today
+date: 5 August 2024
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper investigates the consumer behaviour of an average Ontario household.
+Drawing upon data from Statistics Canada, this paper finds people do not spend less in total, nor in food or shelter.
+Our discovery can help policymakers use this information to employ strategies for the wellbeing of the state and to ensure households and states don’t go deeper into debt."
 format: pdf
@@
 The relationship between these variables can provide information in regards to improving economic stability and effective social governmental assistance to support households in managing their spending in the face of inflationary pressures.
-Overall, policymakers can use this information to employ strategies for the wellbeing of the state and ensure households and even states don’t go deeper into debt.
 
